---
title: "Predicting Barbell-Lift Execution from Wearable Sensors"
author: "Ahmed"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    self_contained: true
    toc: true
    toc_depth: 2
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 7, fig.height = 4.2)
train_url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
test_url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
if (!file.exists("data/pml-training.csv")) {
  dir.create("data", showWarnings = FALSE)
  download.file(train_url, "data/pml-training.csv", mode = "wb")
}
if (!file.exists("data/pml-testing.csv")) {
  download.file(test_url, "data/pml-testing.csv", mode = "wb")
}
if (!file.exists("results/model_results.rds")) source("train_model.R")
res <- readRDS("results/model_results.rds")
train_raw <- read.csv("data/pml-training.csv",
  na.strings = c("NA", "", "#DIV/0!"), check.names = TRUE)
test_raw <- read.csv("data/pml-testing.csv",
  na.strings = c("NA", "", "#DIV/0!"), check.names = TRUE)
```

## Executive summary

I built a bagged classification-tree ensemble to predict the five exercise
classes from belt, arm, forearm, and dumbbell accelerometers. After removing
identifiers, timestamps, window fields, and variables with at least 5% missing
values, 52 sensor predictors remained. Stratified five-fold cross-validation
produced **`r sprintf("%.2f%%", 100 * res$cv_overall)` accuracy**, so the
estimated out-of-sample error is **`r sprintf("%.2f%%", 100 * (1-res$cv_overall))`**.
The final ensemble was trained on all 19,622 observations and generated the 20
quiz predictions listed below.

## Data preparation and exploratory analysis

The response `classe` records correct execution (A) and four specific errors
(B--E). The seven administrative variables were excluded because they describe
the observation or collection window rather than body motion and could cause
spurious participant/time leakage. The original data also contain summary
variables that are almost entirely missing; columns with 5% or more missing
values were removed. The retained predictors are complete in both datasets, so
imputation was unnecessary.

```{r class-balance, echo=FALSE}
class_n <- table(train_raw$classe)
barplot(class_n, col = "#377eb8", border = NA,
        xlab = "Exercise class", ylab = "Training observations",
        main = "Outcome classes are moderately balanced")
```

Class A is most common, but every class has more than 3,000 observations.
Stratification therefore preserves the class distribution in every validation
fold and prevents the accuracy estimate from being dominated by class A.

## Model and cross-validation

Classification trees accommodate nonlinear relationships and interactions
without requiring sensor scaling. A single deep tree is unstable, so I fitted
an ensemble of 35 bootstrap trees per validation fold. Each tree used a random
subset of 35 of the 52 predictors; majority voting reduced variance and supplied
additional diversity. The tree controls were `cp = 0`, `minsplit = 5`,
`minbucket = 2`, and `maxdepth = 30`.

Five stratified folds were created once with seed 20260723. For each fold, the
model was trained on the other four folds and evaluated only on the held-out
observations. This makes every observation an out-of-fold prediction and avoids
using the 20 quiz cases for model selection.

```{r cv-results, echo=FALSE}
barplot(100 * res$cv_accuracy, names.arg = paste("Fold", 1:5),
        ylim = c(98, 100), col = "#4daf4a", border = NA,
        ylab = "Accuracy (%)", main = "Five-fold cross-validation accuracy")
abline(h = 100 * res$cv_overall, lty = 2, col = "#e41a1c", lwd = 2)
```

Fold accuracies ranged from `r sprintf("%.2f%%",100*min(res$cv_accuracy))` to
`r sprintf("%.2f%%",100*max(res$cv_accuracy))`. The pooled confusion matrix is:

```{r confusion}
res$cv_confusion
```

The pooled cross-validated error,
`r sprintf("%.2f%%", 100*(1-res$cv_overall))`, is the expected out-of-sample
error. It is preferable to training accuracy because each prediction was made
by a model that did not see that observation. Remaining error may reflect
similar motion patterns between incorrect techniques and participant-specific
variation.

## Final model and quiz predictions

After validation, I trained 75 bootstrap trees on all training observations
using the same preprocessing and settings. The resulting predictions are:

```{r predictions, echo=FALSE}
prediction_table <- data.frame(
  problem_id = test_raw$problem_id,
  prediction = as.character(res$predictions)
)
knitr::kable(prediction_table, align = c("c", "c"))
```

The comma-separated answer file and 20 one-answer text files are included in
the repository. The complete reusable fitting code is in `train_model.R`.

## Reproducibility

```{r session-info}
sessionInfo()
```

## Data citation

Velloso, E., Bulling, A., Gellersen, H., Ugulino, W., and Fuks, H. (2013).
*Qualitative Activity Recognition of Weight Lifting Exercises*. Proceedings of
the 4th Augmented Human International Conference.
